Research on The Impact of Past Behaviour Normality on Regret

### 2024-02-15 outputs/paper/paper.qmd
completed

---
title: "Research on The Impact of Past Behaviour Normality on Regret"
subtitle: "The replication of two major experiments"
author: 
  - Ziheng Zhong
  - Yihang Cai
thanks: "Code and data are available at: https://github.com/iJustinn/Impact_of_Past_Behavior_Normality_on_Regret.git"
date: today
date-format: long
abstract: "To be written after we finished the Discussion section"
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

@@

We categorize social norms into two main types: descriptive norms, which concern beliefs about common behaviors within a group, and injunctive norms, which relate to beliefs about what behaviors are approved or disapproved by others [@citeUNICEF]. Our research involves two experiments aimed at assessing the effect of these norms on regret. The first experiment looks into the role of these norms in the context of deciding whether or not to pick up a hitchhiker, while the second addresses the consequences of selecting an unconventional driving route.

@@

## Method
In the first experiment, participants were presented with a well-known scenario by Kahneman and Miller, often referred to as the hitchhiker scenario: "Mr. Jones almost never takes hitch-hikers in his car. Yesterday he gave a man a ride and was robbed. Mr. Smith frequently takes hitch-hikers in his car. Yesterday he gave a man a ride and was robbed" (@citeOriginalPaper). The original study included three comprehension questions to confirm participants' understanding of the scenario, which we will not detail here. To assess the influence of injunctive and descriptive social norms, participants were asked: "Whose behavior do you think is more common in society? Mr.Jones(1)/Mr.Smith(2)" and "Whose behavior do you think will be more criticized by others in society? Mr.Jones(1)/Mr.Smith(2)" respectively. And the level of regret is also assessed by asking the question "Who do you expect will experience greater regret over the episode? Mr.Jones(1)/Mr.Smith(2)" (@citeOriginalPaper). The additional measure of negative effect is measured by asking the question "Contemplating your previous answers about this scenario and factoring in both Mr.Jones and Mr.Smith personal routines and your perceptions of social norms and possible social criticism, who do you think overall experienced more negative feelings about the decision to take a hitch-hiker that day? Mr.Jones(1)/Mr.Smith(2)".

@@

## Attibutes
The survey dataset initially contained 52 variables. After excluding variables related to the third experiment and comprehension questions, the dataset was narrowed down to 6 key variables for analysis. For the first experiment, 'Sc1_regret' represents participant responses to the regret question. 'Sc1_socnorms1' and 'Sc1_socnorms2' correspond to the assessment of injunctive and descriptive social norms, respectively. 'Sc1_combinednorms' is an additional variable introduced to measure the overall negative emotional response. As mentioned above, 1 in the dataset represents the Mr.Jones, where 2 represents Mr.Smith.

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: tbl-filtereddata
#| tbl-cap: top 10 rows of filtered data

# Results
## Experiment 1

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: tbl-experiment1
#| tbl-cap: Result of experiment 1

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: fig-experiment1bar
#| fig-cap: Bar plot of experiment 1

@@

# Create the bar plot with the categories reordered
ggplot(data_long, aes(x = Category, y = Percentage, fill = Condition)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  scale_fill_manual(values = c("Smith_Routine" = "grey", "Jones_Exception" = "#505050"), 
                    labels = c("Mr.Smith(Routine)", "Mr.Jones(Exception)")) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  theme_minimal() +
  labs(x = "", y = "Percentage", fill = "Condition") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, colour = "black",size = 9), 
    legend.position = "top", legend.title = element_blank())
```

@tbl-experiment1 and @fig-experiment1bar display results from experiment 1. Mr. Jones, who performs an action that is an exception to his routine, reports a higher regret level of 92.1%. Most participants rated Mr.Jones' behavior to be more common (descriptive) as he never gave ride to hitchhiker (95.3) and therefore to be less criticized (injunctive) by the society (9.4%). Mr.Jones is thought to experience more negative feelings about the decision made that day combining previous answers to the two social norms (92.7%). Mr.Smith, associated with routine behavior, shows markedly lower levels in these aspects, participants rated 7.9%, 4.7%, 90.6% and 7.3% for regret, descriptive social norm, injunctive social norm, and negative effect respectively. Indicating that although Mr.Smith is thought to be criticized more in the society, he would experience less regret and negative effect overall.

The disparity in responses suggests that individuals who engage in behaviors that are not typical for them (Mr. Jones) are more likely to experience and be perceived as having stronger regret and more negative emotional responses. The consistency in this pattern across different measures indicates a robust finding, aligning with psychological theories that predict increased emotional consequences for actions that deviate from one's own or society's expectations. The bar plot @fig-experiment1bar quantitatively illustrates these findings. Mr. Jones's experiences with regret and negative affect are visually represented by significantly taller bars compared to Mr. Smith's, reflecting the numerical data.

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: tbl-experiment2
#| tbl-cap: Result of experiment 2

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'h'
#| label: fig-experiment2bar
#| fig-cap: Bar plot of experiment 2

adams_regret_p <- round((mr_adams_counts / total_counts_2) * 100, 1)
adams_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 1) / total_counts_2) * 100, 1)
white_regret_p <- round((sum(data_filtered_2$Sc2_regret == 2) / total_counts_2) * 100, 1)
white_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 2) / total_counts_2) * 100, 1)

@@

## Replication and estimand
In replicating the classic experiments by Kahneman and Miller (1986), our study sought to precisely estimate the effect of deviations from routine on the experience of regret — a clearly defined estimand. This estimand was articulated in the introduction and guided the experimental design to ensure the validity of our findings.

## Measurement validity
The validity of our measurements was paramount, as we endeavored to capture quantifiable aspects of regret that directly relate to our estimand. Our measures of regret, social norms, and the added dimension of luck were developed to reflect the actual constructs under study, adhering to the principles outlined by Flake and Fried (2020). This rigorous approach to measurement ensures that the quantified variables are not only appropriate but also reliably represent the concepts they are intended to measure.

## Ethical Considerations and Bias
Ethical considerations in replicating behavioral experiments are critical. We ensured the ethical treatment of participants and transparency in our methodology, especially considering the subjective nature of self-reported data on regret. The study's design attempted to minimize bias, but we acknowledge that no research is free from all bias, including potential measurement errors or the influence of the study's context on participants' responses.

## Limitations and Future Research
A limitation of the current study includes the use of vignettes, which may not fully capture the complexity of real-life decisions and their emotional consequences. Future studies might explore the phenomenon in naturalistic settings or employ longitudinal designs to observe the persistence of regret over time. Additional research could also examine whether different types of regret have varying impacts on future decision-making and behavior change.

## Conclusion
Our study contributes to the body of literature on decision-making and emotion by providing a refined estimation of the exceptionality effect's impact on regret. The robustness of the effect across replications reinforces its relevance in psychological research. Moving forward, there is a need for more nuanced investigations that consider the complexity of human behavior and the myriad factors that shape our emotional responses to our decisions.

>>>>

---
title: "Research on The Impact of Past Behaviour Normality on Regret"
subtitle: "The replication of two major experiments"
author: 
  - Ziheng Zhong
  - Yihang Cai
thanks: "Code and data are available at: https://github.com/iJustinn/Impact_of_Past_Behavior_Normality_on_Regret.git, the original paper can be found at https://doi.org/10.1080/02699931.2018.1504747"
date: today
date-format: "long" 
abstract: "The paper investigates the influence of deviation from routine versus social norms on regret. Through replicating two significant experiments, it found that deviations, or the exceptionality effect, significantly impact regret levels. Participants perceived those who diverged from routine behaviors as experiencing more regret, a finding consistent across two experiments and further underscored by measurements of negative effect and luck. This study enriches the understanding of decision-making and emotions, emphasizing the pivotal role of personal behavioral norms in shaping regret, thereby contributing to psychological research's knowledge base."
format: pdf
output: bookdown::pdf_document2
toc: true
number-sections: true
bibliography: references.bib
---

@@

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

@@

We are trying to measure which of the two factors has a more significant effect on the level of regret (estimand): social norms and past behaviors(exceptionality effect). We categorize social norms into two main types: descriptive norms, which concern beliefs about common behaviors within a group, and injunctive norms, which relate to beliefs about what behaviors are approved or disapproved by others [@citeUNICEF]. Our research involves two experiments aimed at assessing the effect of these norms on regret. The first experiment looks into the role of these norms in the context of deciding whether or not to pick up a hitchhiker, while the second addresses the consequences of selecting an unconventional driving route.

@@

## Method

In the first experiment, participants were presented with a well-known scenario by Kahneman and Miller, often referred to as the hitchhiker scenario: "Mr. Jones almost never takes hitch-hikers in his car. Yesterday he gave a man a ride and was robbed. Mr. Smith frequently takes hitch-hikers in his car. Yesterday he gave a man a ride and was robbed" (@citeOriginalPaper). The original study included three comprehension questions to confirm participants' understanding of the scenario, which we will not detail here. To assess the influence of injunctive and descriptive social norms, participants were asked: "Whose behavior do you think is more common in society? Mr.Jones(1)/Mr.Smith(2)" and "Whose behavior do you think will be more criticized by others in society? Mr.Jones(1)/Mr.Smith(2)" respectively. And the level of regret is also assessed by asking the question "Who do you expect will experience greater regret over the episode? Mr.Jones(1)/Mr.Smith(2)" (@citeOriginalPaper). The additional measure of negative effect is measured by asking the question "Contemplating your previous answers about this scenario and factoring in both Mr.Jones and Mr.Smith personal routines and your perceptions of social norms and possible social criticism, who do you think overall experienced more negative feelings about the decision to take a hitch-hiker that day? Mr.Jones(1)/Mr.Smith(2)".

@@

## Attibutes

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-filtereddata
#| tbl-cap: top 10 rows of filtered data

The survey dataset initially contained 52 variables. After excluding variables related to the third experiment and comprehension questions, the dataset was narrowed down to 6 key variables for analysis. For the first experiment, 'Sc1_regret' represents participant responses to the regret question. 'Sc1_socnorms1' and 'Sc1_socnorms2' correspond to the assessment of injunctive and descriptive social norms, respectively. 'Sc1_combinednorms' is an additional variable introduced to measure the overall negative emotional response. As mentioned above, 1 in the dataset represents the Mr.Jones, where 2 represents Mr.Smith.

@@

# Results

## Experiment 1

@tbl-experiment1 displays results from experiment 1. Mr. Jones, who performs an action that is an exception to his routine, reports a higher regret level of 92.1%. Most participants rated Mr.Jones' behavior to be more common (descriptive) as he never gave ride to hitchhiker (95.3) and therefore to be less criticized (injunctive) by the society (9.4%). Mr.Jones is thought to experience more negative feelings about the decision made that day combining previous answers to the two social norms (92.7%). Mr.Smith, associated with routine behavior, shows markedly lower levels in these aspects, participants rated 7.9%, 4.7%, 90.6% and 7.3% for regret, descriptive social norm, injunctive social norm, and negative effect respectively. Indicating that although Mr.Smith is thought to be criticized more in the society, he would experience less regret and negative effect overall.

```{r}
#| message: false
#| echo: false
#| warning: false
#| fig-pos: 'H'
#| label: tbl-experiment1
#| tbl-cap: Result of experiment 1

@@

The disparity in responses suggests that individuals who engage in behaviors that are not typical for them (Mr. Jones) are more likely to experience and be perceived as having stronger regret and more negative emotional responses. The consistency in this pattern across different measures indicates a robust finding, aligning with psychological theories that predict increased emotional consequences for actions that deviate from one's own or society's expectations.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-experiment1bar
#| fig-cap: Bar plot of experiment 1

@@

# Create the bar plot with the categories reordered
ggplot(data_long, aes(x = Category, y = Percentage, fill = Condition)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.7) +
  scale_fill_manual(values = c("Smith_Routine" = "grey", "Jones_Exception" = "#505050"), 
                    labels = c("Mr.Smith(Routine)", "Mr.Jones(Exception)")) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  theme_minimal() +
  labs(x = "", y = "Percentage", fill = "Condition") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 1, colour = "black",size = 9), 
    legend.position = "top", legend.title = element_blank())
```
The bar plot @fig-experiment1bar quantitatively illustrates these findings. Mr. Jones's experiences with regret and negative affect are visually represented by significantly taller bars compared to Mr. Smith's, reflecting the numerical data.

@@

```{r}
#| message: false
#| fig-pos: 'h'
#| echo: false
#| warning: false
#| label: tbl-experiment2
#| tbl-cap: Result of experiment 2

@@

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-experiment2bar
#| fig-cap: Bar plot of experiment 2

# calculate each percentage
adams_regret_p <- round((mr_adams_counts / total_counts_2) * 100, 1)
adams_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 1) / total_counts_2) * 100, 1)
white_regret_p <- round((sum(data_filtered_2$Sc2_regret == 2) / total_counts_2) * 100, 1)
white_luck_p <- round((sum(data_filtered_2$Sc2_lucky == 2) / total_counts_2) * 100, 1)

@@

## Replication and estimand

In replicating the classic experiments by Kahneman and Miller (1986), our study sought to estimate the effect of social norms and past behaviors on the experience of regret (which of the two is more important) — a clearly defined estimand. This estimand was articulated in the introduction and guided the experimental design to ensure the validity of our findings.

## Measurement validity

The validity of our measurements was crucial, as we endeavored to capture quantifiable aspects of regret that directly relate to our estimand. Our measures of regret, social norms, and the added dimension of luck and negative effect were developed to reflect the actual constructs under study. This rigorous approach to measurement ensures that the quantified variables are not only appropriate but also reliably represent the concepts they are intended to measure.

## Ethical Considerations and Bias

Ethical considerations in replicating behavioral experiments are critical. The original paper recruited American Amazon Mechanical Turk (MTurk) participants online, and ensured that the participants took the survey in the first experiment could not take in the second experiment. Thus, eliminating any possible effects between the two experiments. The study's design attempted to minimize bias, but we acknowledge that no research is free from all bias, including potential measurement errors or the influence of the study's context on participants' responses.

## Limitations and Future Research

A limitation of the current study includes the use of vignettes, which may not fully capture the complexity of real-life decisions and their emotional consequences. Future studies might explore the phenomenon in naturalistic settings or employ longitudinal designs to observe the persistence of regret over time. Additional research could also examine whether different types of regret have varying impacts on future decision-making and behavior change.

## Result and Conclusion

The study's findings suggest that deviations from routine have a more pronounced impact on regret than social norms. Both experiments showed that participants believed individuals who diverged from their usual behavior (Mr. Jones and Mr. White) felt more regret compared to those who did not. This was further supported by the negative effect measurement, which took into account both injunctive and descriptive social norms. Additionally, participants considered Mr. White to be less fortunate than Mr. Adams, suggesting that routine actions might lead to fewer regrets. This research adds valuable insight into how exceptional actions influence regret, reinforcing the significance of the exceptionality effect in understanding emotional responses in decision-making. The robustness of the effect across replications reinforces its relevance in psychological research.
